---
title: "SOC6707 Assignment 1 Template"
author: "Your name here"
date: "date here"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

```{r,  echo = FALSE}
library(tidyverse) # need to load in tidyverse
```


# Question 1

## a)

i) 
YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


ii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


iii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


## b)

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

## c)

i) 
YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


ii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


iii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


# Question 2

## a)

i) 
YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


ii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


iii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

## b)

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

## c)

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


# Question 3

## a) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


## b)

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


## c) 


i) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

ii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

iii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


## d) 

i) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```

ii) 

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```


## e)

YOUR ANSWER HERE

```{r}
# YOUR CODE HERE
```




