---
title: "R Basics"
author: "Monica Alexander"
date: "09/01/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R basics

## Assign values to variables

The chunk above we used R as a simple calculator (2+2) We can also assign **values** to **variables** with the back arrow i.e. `<-`. For example (execute this chunk to see the outcomes)

```{r}
# assigning the variable x to have a value of 1
x <- 1
# assigning the variable y to have a value of 2
y <- 2
# print these
x
y
# we can add these together too
x+y
```

Side note: you may see in other R codes that `=` is also used to assign values to a variable. 

Values need not just be numbers:

```{r}
# the c() function allows you to create vectors of numbers (or characters)
z <- c(3,4,3.2,5.1)
z
# pull out variable parts of z
z[1]
z[3]
# length of z
length(z)
# the 5th element doesn't exist
z[5]
```

Character strings:

```{r}
instructor_name <- "Monica Alexander"
first_name <- "Monica"
last_name <- "Alexander"
```

**Functions** in R are commands that take arguments and do operations to variables/objects. For example, the `paste` function pastes two (or more) strings together:

```{r}
z
max(z)
mean(z)
```


```{r}
paste(first_name, last_name)
paste(first_name, "June", last_name)
```

*Sidenote*: R is sensitive to capitalization, both in commands and in variable names. For example using `Paste` you would get an error. 

## Practice Exercise 

Using the codes you learned, write a code that outputs, "your name, program of study". 


## Getting help

To see what a function does, and to check the arguments, type a "?" and then the function name, for example:

```{r}
?paste
```

Once you execute the code above, you should see that the help file for paste has appeared in the bottom right pane.

## Logical statements

It is useful to check to see if variables or objects are less than, equal to or greater than numbers. Below are some examples. Note that:

- Equality is two = signs (not one)
- Each of these statements returns a **logical** value i.e. TRUE or FALSE

```{r}
#equals
x==1
x==2
# greater than
y>1
x>1
# greater than or equal to
x>=1
# less than
x<9
```

