---
title: 'Week 10:  Multinomial regression'
author: "Monica Alexander"
date: "18/03/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


Packages:

```{r}
library(tidyverse)
library(here)
library(nnet) # for multinomial
```


# Multinomial

Question of interest: how does infant mortality cause of death vary by race, mother's age and prematurity?

## Data prep

Read in infant data and do some cleaning:

```{r}
d <- read_rds(here("data/infant.RDS"))
d <- d %>% 
  mutate(neo_death = ifelse(aged<=28, 1, 0),
         cod_group = case_when(
           str_starts(cod, "peri") ~ "perinatal",
           cod %in% c("other", "unknown") ~ "other/unknown",
           cod %in% c("sids", "maltreatment", "infection") ~ "exogenous",
           cod %in% c("resp", "heart") ~ "respiratory/heart",
           TRUE ~ "congenital malformations"
         ),
         preterm = ifelse(gest<37, 1, 0)) %>% 
  filter(gest<99, !is.na(mom_age_group))

infant <- d %>% select(race, mom_age, gest, preterm, cod_group)
infant
```

Graph by race from lecture:

```{r}
infant %>% 
  group_by(race, cod_group) %>%
  tally() %>% 
  group_by(race) %>% 
  mutate(prop = n/sum(n)) %>%
  mutate(cod_group = fct_reorder(cod_group, -prop)) %>% 
  ggplot(aes(cod_group, prop, fill  = race)) + geom_bar(stat = "identity", position = 'dodge') + 
  labs(title = "Proportion of infant deaths by cause", x = "cause", y = "proportion") + 
  theme_bw() + 
  scale_fill_brewer(palette = "Set1")
```

Making into wide format

```{r}
infant_wide <- infant %>% 
  group_by(race, mom_age, gest, preterm, cod_group) %>% 
  tally(name = "deaths") %>% 
  pivot_wider(names_from = cod_group, values_from = deaths) %>% 
  mutate_all(.funs = funs(ifelse(is.na(.), 0, .)))
head(infant_wide)
```

Making the Y variable:

```{r}
infant_wide$Y <- as.matrix(infant_wide[,c("perinatal",
                                          "exogenous",
                                          "congenital malformations", 
                                          "respiratory/heart", "other/unknown")])
head(infant_wide$Y)
```

## Regression

```{r}
library(nnet)
mod_mn <- multinom(Y ~ race+ mom_age+ preterm, data = infant_wide)
summary(mod_mn)
```

Pull out coefficients:

```{r}
coef(mod_mn)
exp(coef(mod_mn))
```

Exercise: plot coefficient estimates and standard errors.


## Predicted probabilities

```{r}
predict_df <- tibble(race = rep(c("NHW", "NHB"), each = 2),
       mom_age = 30,
       preterm = rep(c(0,1),2)) 
preds <- bind_cols(predict_df, as_tibble(predict(mod_mn, newdata = predict_df, type = 'probs')))
preds
```

Plot:

```{r}
preds %>% 
  pivot_longer(`perinatal`:`other/unknown`, names_to = "cod_group", values_to = "probability") %>% 
  mutate(preterm = ifelse(preterm==1, "pre-term", "full-term")) %>% 
  ggplot(aes(race, probability, fill = cod_group)) + 
  geom_bar(stat = "identity")+
  facet_grid(~preterm) +
  ggtitle("Predicted probabilities of infant death by race, prematurity and cause\nMothers aged 30")
```










