---
title: "Example RMarkdown"
author: "Monica Alexander"
date: "10/01/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hi my name is Monica blah blah 

```{r}
# this is a comment
x <- 853
y <- 90
x+y
```

Read in the GSS dataset

```{r}
#  load in tidyverse package 
library(tidyverse)
# read in the GSS
gss <- read_csv(file = "../data/gss.csv")
```

Look at the top of the GSS dataset

```{r}
head(gss)
```
Look at the dimensions of the GSS

```{r}
ncol(gss) # number of columns
nrow(gss) # rows
```

# Important functions

## Select columns

Let's look at column names. The function colnames returns a vector of length 85, showing us every column name

```{r}
# column names
colnames(gss)
```

Let's select the age variable

```{r}
select(gss,age)
```

## The pipe 

The pipe is the funny thing that is %>% -- remember from the lecture, when you see a pipe, read it as "and then"

```{r}
gss %>% 
  select(age, region, occupation)
```
NOTE: the above code selects three columns, but doesn't assign/save the output. If we want to assign this output to use again, just use the back arrow syntax

```{r}
gss_ageregionocc <- gss %>% 
  select(age, region, occupation)

# look at this object
gss_ageregionocc
```

```{r}

```

















